\mbox{Verifying Imperative Higher-Order Programs}

### 2009-10-06 papers/esop2010/main.tex

\author{Johannes Kanig\inst{1,}\inst{2} \and Jean-Christophe
Fill\^atre\inst{2,}\inst{1}}
\institute{
INRIA Saclay - Île-de-France, ProVal, Orsay, F-91893 \and
 LRI, Univ Paris-Sud, CNRS, Orsay, F-91405
}
%TODO explain somewhere the correspondence store type <-> map, effect variable
%<-> domain, region <-> key
%TODO the constants dealing with maps

%TODO clearly state possibilities/limitations of the system
% * one can store everything in refs
% * but aliasing restriction (refs in refs, lists of refs etc)

@@

\begin{figure}[tbp]
  \begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~[\alist{κ}] :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ, x : τ' \\ Γ'' = Γ',~f : τ' ->^ε τ \\ Γ'' |- t : τ,ε \\ Γ' |-_l
  p : \efft{ε} -> \propml \\ Γ' |-_l q : \efft{ε} ->
  \efft{ε} -> τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v t1 : τ' -> τ \\ Γ|-_v t2 : τ'}
    {Γ |-_v  t1 ~ t2 : τ}
  }
  \end{mathpar}
  \caption{Value typing}
  \label{fig:valuetyping}
\end{figure}

@@

\begin{figure}[tpb]
\begin{mathpar}
  { \inferrule*[Left=Value]
  { Γ |-_v t : τ  }
  { Γ |- t : τ,\emptyset  }
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v t1 : τ' ->^ε τ \\ Γ|-_v t2 : τ'}
    {Γ |- t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |- t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- t1 : τ', ε \\ Γ, x : τ' |- t2 : τ, ε }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε }
  } \and
  { \inferrule*[Left=Sub]
    { Γ |- t : τ', ε \\ ε \subseteq ε' }
    { Γ |- t : τ, ε' }
  }
\end{mathpar}
  \caption{Program typing}
  \label{fig:progtyping}
\end{figure}

@@

\begin{figure}[tbp]
  \begin{mathpar}
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  } \qquad \and
  { \inferrule*[Left=LetLogic]
    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |-_l t : τ  }
    { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
  }\and
  { \inferrule*[Left=IteLogic]
    { Γ |- _l t1 : \boolml \\ Γ |-_l t2 : τ \\ Γ |-_l t3 : τ }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ }
  }
  \end{mathpar}
  \caption{Logic typing}
  \label{fig:logictyping}
\end{figure}

@@

The presented programming language, along with the specifications, can, in
fact be used to prove programs. The weakest precondition calculus presented in
the next section can be used to obtain proof obligations which, when proved,
guarantee the correctness of the program with respect to its specification. On
the other hand, in its presented form, and because of the presence of
aliasing, one would constantly need equalities and disequalities between
reference variables to be able to reason about a program.

A way out is to restrict programs to alias free programs. Fortunately, it is
easy to modify the typing rules to render all programs with aliasing
ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
in such a way that there is exactly one memory location in each region.
Starting from this assumption, a program can actually
exhibit {\em two} forms of aliasing. In the first one, it may use several
names (region variables) for the same memory location. This can happen when a
function is generalized over, say, two region variables, assuming they are
different, and is then instantiated with the same region variable twice. The
classical example is the following program
\begin{who}
  let f [ρ ρ'] (x : int ref(ρ)) (y : int ref(ρ')) =
  { }
  x := !x + 1;
  !y
  { r : r = !!y old }
\end{who}
which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
this, we add a side condition to the instantiation: the substitution $σ$ must
be {\em compatible} with the type $τ$, written $σ\sim τ$. To understand what
compatibility means, consider an {\em atomic} region substitution (whose
domain contains only one variable) $[ρ|->ρ']$. It is compatible with any type
which does not contain $ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2
e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
Expressed more generally, an atomic substitution is compatible with a type if
the type does not contain any region or effect variables which are in the
range of the substitution. Finally, a {\em composite} substitution (whose
domain contains more than one variable) is compatible with a type if each
atomic substitution is compatible with the result of the application of the
previous substitution. In this way, it becomes impossible to have different
names for a single reference cell.

\subsection{Some properties}

%TODO
$Γ|-_v v : τ$ implies $Γ |-_l \ceil{v} : \ceil{τ}$

Subject reduction of our type system is a consequence of subject reduction of
the region calculus of Tofte/Talpin~\cite{tofte97ic}.
%TODO ref vers l'annexe -  remarque sur la difference

@@

%TODO lemmas about correctness
\section{Examples}
\label{sec:examples}

>>>>

\author{Johannes Kanig\inst{1,}\inst{2} \and Jean-Christophe
Fill\^atre\inst{2,}\inst{1}}
\institute{
INRIA Saclay - Île-de-France, ProVal, Orsay, F-91893 \and
 LRI, Univ Paris-Sud, CNRS, Orsay, F-91405
}
%TODO explain somewhere the correspondence store type <-> map, effect variable
%<-> domain, region <-> key

@@

\begin{figure}[tbp]
  \framebox{$Γ|-_v t : τ$}
  \begin{mathpar}
  { \inferrule*[Left=Var] {Γ(x) = ∀\alist{χ}.τ} {Γ|-_v x~[\alist{κ}] :
  τ[\alist{χ}|->\alist{κ}]} } \and
  { \inferrule*[Left=PureFun] {Γ,x : τ' |-_v t : τ} {Γ|-_v λx : τ'.t : τ' ->
  τ} } \and
  { \inferrule*[Left=Rec] 
  { Γ' = Γ, x : τ' \\ Γ'' = Γ',~f : τ' ->^ε τ \\ Γ'' |- t : τ,ε \\ Γ' |-_l
  p : \efft{ε} -> \propml \\ Γ' |-_l q : \efft{ε} ->
  \efft{ε} -> τ -> \propml } 
      {Γ|-_v \recml~f~(x : τ').\{p\}t\{q\} : τ' ->^ε τ} 
  } \and
  { \inferrule*[Left=PureApp]
    {Γ |-_v t1 : τ' -> τ \\ Γ|-_v t2 : τ'}
    {Γ |-_v  t1 ~ t2 : τ}
  }
  \end{mathpar}
  \caption{Value typing}
  \label{fig:valuetyping}
\end{figure}

@@

\begin{figure}[tpb]
  \framebox{$Γ|- t : τ,ε$}
\begin{mathpar}
  { \inferrule*[Left=Value]
  { Γ |-_v t : τ  }
  { Γ |- t : τ,\emptyset  }
  } \and
  { \inferrule*[Left=App]
    {Γ |-_v t1 : τ' ->^ε τ \\ Γ|-_v t2 : τ'}
    {Γ |- t1 ~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=ite]
    { Γ |- _v t1 : \boolml \\ Γ |- t2 : τ, ε1 \\ Γ |- t3 : τ, ε2 }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ, ε1  ε2 }
  } \and
  { \inferrule*[Left=Letv]
    { Γ, \alist{χ} |-_v t1 : τ' \\ Γ, x : ∀\alist{χ}.τ' |- t2 : τ,ε }
    { Γ |- \letml~ x~[\alist{χ}] = t1 ~ \inml~ t2 : τ, ε}
  } \and
  { \inferrule*[Left=Let]
    { Γ |- t1 : τ', ε \\ Γ, x : τ' |- t2 : τ, ε }
    { Γ |- \letml~ x = t1 ~ \inml~ t2 : τ, ε }
  } \and
  { \inferrule*[Left=Sub]
    { Γ |- t : τ', ε \\ ε \subseteq ε' }
    { Γ |- t : τ, ε' }
  }
\end{mathpar}
  \caption{Program typing}
  \label{fig:progtyping}
\end{figure}

@@

\begin{figure}[tbp]
  \framebox{$Γ|-_l t : τ$}
  \begin{mathpar}
  { \inferrule*[Left=Logic]
    { Γ |-_v t : τ  }
    { Γ |-_l t : \ceil{τ} }
  } \qquad \and
  { \inferrule*[Left=LetLogic]
    { Γ, \alist{χ} |-_l : t' : τ' \\ Γ, x: ∀\alist{χ}.τ' |-_l t : τ  }
    { Γ |-_l \letml~x~[\alist{χ}] = t'~\inml~t : τ }
  }\and
  { \inferrule*[Left=IteLogic]
    { Γ |- _l t1 : \boolml \\ Γ |-_l t2 : τ \\ Γ |-_l t3 : τ }
    { Γ |- \ifml~ t1 ~\thenml~ t2 ~\elseml ~ t3 : τ }
  }
  \end{mathpar}
  \caption{Logic typing}
  \label{fig:logictyping}
\end{figure}

@@

@@

The presented programming language, together with the weakest precondition
calculus, form a basis for program verification and can already be used to
prove programs. On the other hand, in its presented form, and because of the
presence of aliasing, one would constantly need equalities and disequalities
between reference variables to be able to reason about a program.

A way out is to restrict the attention to alias free programs. Fortunately, it
is easy to modify the typing rules to render all programs with aliasing
ill-typed (see Fig~\ref{fig:aliasing}). In the following, we will use regions
in such a way that there is exactly one memory location in each region.
Starting from this assumption, a program can actually exhibit {\em two} forms
of aliasing. In the first one, it may use several names (region variables) for
the same memory location. This can happen when a function is generalized over,
say, two region variables, assuming they are different, and is then
instantiated with the same region variable twice. The classical example is the
following program
\begin{who}
  let f [ρ ρ'] (x : int ref(ρ)) (y : int ref(ρ')) =
  { }
  x := !x + 1;
  !y
  { r : r = !!y old }
\end{who}
which is only correct if $ ρ $ and $ ρ' $ are indeed different. To enforce
this, we add a side condition to the instantiation: the substitution $σ$ must
be {\em compatible} with the type $τ$, written $σ\sim τ$. To understand what
compatibility means, consider an {\em atomic} region substitution (whose
domain contains only one variable) $[ρ|->ρ']$. It is compatible with any type
which does not contain $ρ'$. Similarly, the effect substitution $[e|-> ρ1  ρ2
e'$] is compatible with any type which does not contain $ ρ1 , ρ2 $ nor $e'$.
Expressed more generally, an atomic substitution is compatible with a type if
the type does not contain any region or effect variables which are in the
range of the substitution. Finally, a {\em composite} substitution (whose
domain contains more than one variable) is compatible with a type if each
atomic substitution is compatible with the result of the application of the
previous substitution. In this way, it becomes impossible to have different
names for a single reference cell.

With the aliasing restriction, complex programs using shared mutable data
structures are impossible to prove. Programs with references in lists,
references in references and so on can be written, but only with the severe
limitation that a list can only contain a single reference (potentially
several times) and that a reference can only contain always the same
(statically known) reference. However, references still are first-class
values, i.e. they can be passed to functions, returned by functions and stored
in data structures, with the above limitations.

\paragraph{Correctness}

In the 
%TODO lemmas about correctness
\section{Examples}
\label{sec:examples}
